Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-03-04 paper.tex
Text improvements

The plasma trough is the transition region between the high density and low temperature plasmasphere and the lower density and, in the equatorial plane, higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Although there are well--validated models of the plasmasphere density as a function of $L$ (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}) and plasmapause boundaries (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}), the dependence of the plasma trough on solar wind conditions has not been extensively studied.

\cite{Gallagher1988EmpiricalModelPlasmasphere} used the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model the the plasmasphere as a simple linear fit dependent on $L$ shell, geographic altitude, geomagnetic latitude, and $K_p$.  They found they could accurately model geomagnetically active conditions by using a weighted average $K_p$ value and an azimuthal symmetry at smaller L-shells.  \cite{Hamilton1988RingCurrentDevelopment} analyzed a strong geomagnetic storm and compared energy measurements of the ring current to $D_{st}$, along with measurements of ring current compositions before, during, and after the storm. They found $D_{st}$ to be correlated with the inner ring current energy density during and after storm onset, an inward movement of energy density, and a rapid $D_{st}$ recovery from the quick loss of heavier ions.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 and found that the average ion mass, $M=\rho/n_e$, correlated with geomagnetic activity, with more negative $D_{st}$ values corresponding to slightly higher average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$, with an increasing correlation as $K_p$ was smoothed to longer timescales.  The CRRES measurements used occurred in the MLT range of 12:00 and 18:00 mostly in the plasma trough (between 2-7$R_E$) on an elliptical, nearly equatorial orbit.

\cite{Denton2006} showed a relationship between $D_{st}$ and the field line distribution of plasma density, and specifically along the same field lines as looked at in later papers (6-8$R_E$). They found that the distribution depends on $L$,.

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the GOES satellites from 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alf\v'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

@@

In this work we consider the dependence of plasmaspheric mass density in the \cite{Takahashi2010} in more detail in an attempt to identify the solar wind processes that drives changes in the plasma trough density.  There are several issues that are addressed in detail: (1) the plasma density measurements are sparse and (2) the dependence on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, the plasmasphere can expand past geosynchronous orbit where measurements of density are made resulting in an increase in observed density; for southward IMF, events studies have shown an increase in density associated with strong southward IMF [\cite{Takahashi2010}; \cite{Yao2008}].  Issue (1) is addressed by considering \note{Finish}.  Issue (2) is addressed by considering epochs and linear correlations independent of the direction of the IMF and also by separation by averages of the northrward and southward IMF.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton}, with data available from 1980 through 1991; all other parameters are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements \vinote{More explanation?}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple Geostationary Operational Environmental Satellites (GOES) satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

@@

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of -50~nT.  The second is an increase in $\rho_{eq}$ above 40~amu/cm$^3$.

@@

Our first analysis uses data only from the time interval 1989-1991 to correspond with the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter were computed using the standard deviation of the values used in computing the median divided by the square root of the number of values used.

In the upper panel of Figure \ref{DailyAveragedDstEvents}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $\sim 2$ amu/cm$^3$ instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events which is much larger than the number of \req\ values as we use all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed. \vinote{Alternate: Note that for $D_{st}$ events, we computed median values using all available data, even when \req\ availability was sparse.}

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire run of GOES 6 over 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

@@

In the lower panel of \ref{DailyAveragedDstEvents}, $\rho_{eq}$ events are shown.  Here we see that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancments to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours. Figure \ref{HourlyAveragedDstEvents} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{HourlyAveragedDstEvents}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-50$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that went into the average. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dd12.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events in the interval 1983-1991; compare to bottom panel {DailyAveragedDstEvents}. Bottom: Same as Top except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours.}
\label{HourlyAveragedDstEvents}
\end{figure}

>>>>

The plasma trough is the transition region between the high density and low temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapuase, the density has contributions from the co-rotating plasmasphere and conveting magnetosphere plasmas.

Although there are well--validated models, as a function of $L$, $MLT$, and geomagnetic activity, of the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}) and plasmapause boundaries (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}), the dependence of the plasma trough on geomagnetic and solar wind conditions has not been extensively studied until the past decade.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 and found that the average ion mass, $M=\rho/n_e$, correlated with geomagnetic activity, with more negative $D_{st}$ values corresponding to slightly higher average ion mass densities.  They also found a relationship between the average ion mass and 1.5- and 3-day averages of $K_p$, with an increasing correlation as $K_p$ was smoothed to longer timescales.  The CRRES measurements used were made in the MLT range of 12:00 and 18:00 and mostly in the plasma trough (between 2-7$R_E$) on an elliptical, nearly equatorial orbit.

\cite{Denton2006} showed a relationship between $D_{st}$ and the field line distribution of plasma density, and specifically along the same field lines as looked at in later papers (6-8$R_E$). They found that the distribution depends on $L$, \note{finish}.

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.

@@

In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify the solar wind processes that drive changes in the plasma trough density.  There are several issues that are addressed in detail: (1) the plasma density measurements are sparse and (2) the dependence on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF and low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where measurements of density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in density associated with strong southward IMF and enhanced magnetospheric activity [\cite{Takahashi2010}; \cite{Yao2008}].  Issue (1) is addressed by considering \note{Finish}.  Issue (2) is addressed by considering epochs and linear correlations independent of the direction of the IMF and also by separation by averages of the northrward and southward IMF.

--

The dependence of the plasmapause location on geomagnetic activity was observed as early as \note{Cite}.  \cite{Gallagher1988EmpiricalModelPlasmasphere} used measurements from the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model the plasmasphere density with a relationship dependent on $L$ shell, altitude, geomagnetic latitude, and observations grouped by a weighted average of $K_p$.  The developed statistical model showed significant improvements when a weighted average of $K_p$ over three days was used to group the observations used in the model.

\cite{Hamilton1988RingCurrentDevelopment} analyzed a strong geomagnetic storm and compared energy measurements of the ring current to $D_{st}$, along with measurements of ring current compositions before, during, and after the storm. They found $D_{st}$ to be correlated with the inner ring current energy density during and after storm onset, an inward movement of energy density, and a rapid $D_{st}$ recovery from the quick loss of heavier ions.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset associated with \cite{Denton} and used in \cite{Takahashi2010}, with data available from 1980 through 1991; all other parameters are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements \vinote{More explanation?}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hourly range was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time in the \cite{Denton} dataset, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index \vnote{Cite Kyoto} below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward and the second is a possible increase in density due to enhanced earthward magnetospheric convection.  The movement of the plasmapause boundary has been well documented and parameterized, most recently by \cite{Moldwin2002ModelPlasmapause} and \cite{OBrien2003EmpiricalPlasmapause}.

The second is an increase in $\rho_{eq}$ above a threshold of 40~amu/cm$^3$.  It is expected that these increases will typically occur after period of quiet geomagnetic activity that follow a geomagnetic storm.  In this case, the plasma pause boundary may move outward past the location of measurements and the increase will be due to measurements being made inside the plasmapause.  A second possible reason for the increase could be due to the above-mentioned increase in magnetospheric convention.

@@

Our first analysis uses data only from the time interval 1989-1991 that corresponds with the interval used in the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered on a window of $\pm 12$ hours of the epoch day was computed, and these averaging windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter were computed using the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for these events.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the value of $\sim 10$ amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we use all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed. \vinote{Alternate: Note that for $D_{st}$ events, we computed median values using all available data, even when \req\ availability was sparse.}

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that all trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, from 1983-1991, including the magnitude of change in $\rho_{eq}$ across event onset.

@@

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statictically significant change in $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slighly elevated at the time of the threhshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.

The bottom panel of \ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.

and the event study of \cite{Yao2008} which showed ...

The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{paperfigures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{paperfigures/stormavs-dd12.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events in the interval 1983-1991; compare to bottom panel {DailyAveragedDstEvents}. Bottom: Same as Top except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.}
\label{HourlyAveragedDstEvents}
\end{figure}
